\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-05-27 paper/subdocuments/data_and_methods.Rnw
title change
new Bib-Library
fixed double quotes and R-paths so paper.Rnw compiles

\section{Data, Measurement concepts,and Methods}

@@

\includegraphics[scale=0.5]{"/figure/income_definition.png}

@@

% Die Deductions sind noch nicht gänzlich sauber abgebildet. Auf der Bilanz-Seite steht, vom steuerbaren Einkommen kommt man, indem vom Reineinkommen die Sozialabzüge(Kinder- und Fremdbetreung) sowie der Abzug für unterstützungsbedürftige Personen abzogen werden. 
% Auf den ESTV-Tabellen sind jedoch noch mehr Abzüge abgebildet. Was soll/muss/können wir nun abziehen? Bin verwirrt...

@@

% Streichen
=======
% 
>>>>>>> d21c323... completed subsection "concepts on measuring economic resources" and "defining income"

@@

\textbf{Summary of tax data drawbacks}
\begin{itemize}
\item Misreporting of incomes (high earners have an interest in getting tax beneficial "income" see above)
\item with aggregated data the adjustment for household members is not possible (no equivalization possible). This is a point, which is hardly mentioned in the "top-income" literature, albeit it is a crucial issue in "theorized" concepts of income. To generalize one can say, that the concept of tax units (individuals and couples + no direct information about household members at least not in the aggregated tax tables) is not congruent with the concept of household. This might influence the overall inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing. What are the assumptions?
\item the tax data income definition summarize labor income, capital income and taxable transfer payments (no means-tested benefits as welfare aid). Distinction between the different income sources is not possible, albeit the mechanisms of changes in the specific distribution can be very different.
\item Some tax units are not included in the tax tables (see Foellmi and Martinez 2013:11f).(a) only taxed cases are included (partly a problem, for certain periodes we know how many cases are excluded because of low or none income)	- (b) cases tax at source (Quellenbesteuerung). Foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. In some border-cantons the share of this group is very high (20\%) 
%Can we ignore this)
(c) staff of international organizations based in Switzerland	- (d) non-fillers show up in the tables as long as they are registred. This persons get an imputed income (older tax tax return and information given by employers. Not registered non-	fillers are not in the records.
- (e) tax evasion. Feld and Frey (2007) report about tax evasion in Switzerland, it should be somewhat above 20 percent on average. With the (strong) assumption that the pattern of tax evasion over time is stable, this is a minor problem for inequaly measures over time.
\end{itemize}

@@

It is well recognized that the traditional sources of wealth distribution data are unlikely to provide an accurate picture of wealth ownership in the top-tail of the distribution. Credit Suisse (2011) makes use of the information in the "Rich Lists" published by Forbes Magazine to adjust the wealth distribution pattern in the highest wealth ranges. \\

@@

\textbf{deductions} Del et al. (2007:477):" we can check with statistics for 1971-72 (as well as later years) presented both by size of income before deductions and income after deductions that adding back deductions does not introduce any significant error in our estimates."
Gorgas and Schaltenegger (2011:5): "..., information on [...] deductions is provided in the tax statistics, thus, we could add the personal deductions to the income data to obtain a consistent series over time". Können wir das auch? Zumindest für gewisse Zeiträume? Das wäre noch gut. \\

Studies on income try to focus on the disposable income, which subtracts certain expenditures from the primary income. Deductions reflect somehow compulsory expenditures and thus taxable income can be seen as a sort of pseudo disposable income. On the other hand deductions can affect the distribution. There are recent studies about the correlation of progressivity and deductions in Switzerland, which examines if deductions have a "perverse redistribution" effect by redistributing income from the lower middle class to the upper middle class (vgl. Peters 2011 and Interpellation Barbara Gysel (2009).

@@

\subsubsection{Estimating the bias}
For most of the observed range (1941 to 2010) we do not have any information how many tax units fall into the category of having income that is not zero but is too little to qualify for federal taxation (lets call those "zeros" for convenience) . However starting 1995, the FTA provides exactly this information for each canton. This enables us to estimate the bias we introduce for each canton and each period between 1995 and 2010. Consequently we can obtain information whether the bias is stable over time (which makes it possible to safely interpret the changes of inequality over time) and whether the bias is different for each canton. \textcolor{red}{Unterschiede zwischen Kantonen wären gut um zu argumentieren, dass andere Länder auch davon betroffen sind, in etwa sowas wie "je höher der Steuerfreibetrag, umso stärker der Bias". Länder de erst sehr spät besteuern (und über nicht Besteuerte dann auch nicht Buch führen) haben einen krassen Bias. Wir könnten dann empfehlungen geben, ab welchem Perzentil man save interpretieren kann oder so.}

One can take a first look at the descriptives plotting the share of zeros over time seperately for each canton .

<<setup>>=
library(foreign)
df <- read.dta("../../data/ginis_und_perzentile.dta")
df$zeros <- df$null_norm/df$cpop # i only use normal cases here
@

<<zero_descriptives>>=
library(ggplot2)
ggplot(df[df$steuerperiode>=1995,], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)
ggplot(df[df$steuerperiode>=1995,], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)+geom_line(aes(y=G_steink),color="blue")
@

@@

\begin{enumerate}
\item There is a small overall upward trend which we assume to be the Federal Administrations inflation adjustments to the tax threshold.
\item Geneva and Tessin show wild changes but those might be explained by the tax gap ("Bemessungslücke") that people exploited when the cantons changed the tax system. It remains unclear however why we can't see similar patter within other cantons.
\item There is some variance and we see different patterns over time and cantons. When estimating gini coefficients or the like we must therefore assume that ignoring "the zeros" leads to a bias that is not stable over time.
\end{enumerate}

@@

By adding the zeros as a seperate group we face several problems. With the exception of Geneva 1995/96 Pareto interpolation of percentiles p20 and above seems viable. Low percentiles however would need to be extrapolated so we might impose unrealistic assuptioms. Even when estimating "safe" measures above p20 (better p50) or a gini coefficient we need to make an assumption about the income structure of that group (a distribution, a mean income or zero-income). To be consistent with the measure of "taxable income" we could assume an income of zero for that group, calculate gini coefficients and compare them with the original (uncorrected/plain) version. To get a rough number we could (cantonwise) check the squared correlation between plain and corrected gini.

@@

The model outputs a test statisic for each canton that tells us whether the variation of the zero-rate over time leads to a significant deviation from the typical "canton gini-level". As the model has a decent fit we are not in great danger of omitted variable bias. Using a joint F-Test we can now test if all canton#zeros interactions are zero.

<<ftest>>=
library(survey)
regTermTest(fit,"kanton:zeros")
@

In our case we can clearly reject the hypothesis that all interactions are zero. This leads to the conclusion  that gini coefficients are biased by the variation of the zero-share which is kind of obvious but at the same time we can use the model to report adjusted gini coefficients. For example one might be interested in how inequality would had developed if the zero-share would have been constant over time. (Note RF: predict all data point using canton, time and the initial OR final zero-share to homogenize the time series)
Furthermore we can quantify how large the bias is and we can do this seperately for tax periods or seperately for cantons.

@@

What can we conclude from that analysis? First one must notice that aggregate measure like the gini (or others) do not always react in the same way when we cut off one part of the distribution, therefore the measures calculated from tax data is biased. On the other hand, the model coefficient of Switzerland as a whole is not significant suggesting that the cantonal biases cancel out each other. This seems plausible. Most of the "tax optimization" happens within Switzerland so the rich people who moved to Schwyz are now missing at another canton.

>>>>

\section{Data, Measurement concepts and Methods}

@@

\includegraphics[width=\maxwidth]{figure/income_definition}

@@

% Die Deductions sind noch nicht gänzlich sauber abgebildet. Auf der Bilanz-Seite steht, zum steuerbaren Einkommen kommt man, indem vom Reineinkommen die Sozialabzüge(Kinder- und Fremdbetreung) sowie der Abzug für unterstützungsbedürftige Personen abzogen werden. 
% Auf den ESTV-Tabellen sind jedoch noch mehr Abzüge abgebildet. Was soll/muss/können wir nun abziehen? Bin verwirrt...

@@

% Streichen
%=======
% 
%>>>>>>> d21c323... completed subsection "concepts on measuring economic resources" and "defining income"

@@

\textbf{Summary of tax data drawbacks}
\begin{itemize}
\item Misreporting of incomes (high earners have an interest in getting tax beneficial ``income'' see above)
\item with aggregated data the adjustment for household members is not possible (no equivalization possible). This is a point, which is hardly mentioned in the ``top-income'' literature, albeit it is a crucial issue in ``theorized'' concepts of income. To generalize one can say, that the concept of tax units (individuals and couples + no direct information about household members at least not in the aggregated tax tables) is not congruent with the concept of household. This might influence the overall inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing. What are the assumptions?
\item the tax data income definition summarize labor income, capital income and taxable transfer payments (no means-tested benefits as welfare aid). Distinction between the different income sources is not possible, albeit the mechanisms of changes in the specific distribution can be very different.
\item Some tax units are not included in the tax tables (see Foellmi and Martinez 2013:11f).(a) only taxed cases are included (partly a problem, for certain periodes we know how many cases are excluded because of low or none income)	- (b) cases tax at source (Quellenbesteuerung). Foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. In some border-cantons the share of this group is very high (20\%) 
%Can we ignore this)
(c) staff of international organizations based in Switzerland	- (d) non-fillers show up in the tables as long as they are registred. This persons get an imputed income (older tax tax return and information given by employers. Not registered non-	fillers are not in the records.
- (e) tax evasion. Feld and Frey (2007) report about tax evasion in Switzerland, it should be somewhat above 20 percent on average. With the (strong) assumption that the pattern of tax evasion over time is stable, this is a minor problem for inequaly measures over time.
\end{itemize}

@@

It is well recognized that the traditional sources of wealth distribution data are unlikely to provide an accurate picture of wealth ownership in the top-tail of the distribution. Credit Suisse (2011) makes use of the information in the ``Rich Lists'' published by Forbes Magazine to adjust the wealth distribution pattern in the highest wealth ranges. \\

@@

\textbf{deductions} Del et al. (2007:477):" we can check with statistics for 1971-72 (as well as later years) presented both by size of income before deductions and income after deductions that adding back deductions does not introduce any significant error in our estimates."
Gorgas and Schaltenegger (2011:5): ``..., information on [...] deductions is provided in the tax statistics, thus, we could add the personal deductions to the income data to obtain a consistent series over time''. Können wir das auch? Zumindest für gewisse Zeiträume? Das wäre noch gut. \\

Studies on income try to focus on the disposable income, which subtracts certain expenditures from the primary income. Deductions reflect somehow compulsory expenditures and thus taxable income can be seen as a sort of pseudo disposable income. On the other hand deductions can affect the distribution. There are recent studies about the correlation of progressivity and deductions in Switzerland, which examines if deductions have a ``perverse redistribution'' effect by redistributing income from the lower middle class to the upper middle class (vgl. Peters 2011 and Interpellation Barbara Gysel (2009).

@@

\subsubsection{Estimating the bias}
For most of the observed range (1941 to 2010) we do not have any information how many tax units fall into the category of having income that is not zero but is too little to qualify for federal taxation (lets call those ``zeros'' for convenience) . However starting 1995, the FTA provides exactly this information for each canton. This enables us to estimate the bias we introduce for each canton and each period between 1995 and 2010. Consequently we can obtain information whether the bias is stable over time (which makes it possible to safely interpret the changes of inequality over time) and whether the bias is different for each canton. \textcolor{red}{Unterschiede zwischen Kantonen wären gut um zu argumentieren, dass andere Länder auch davon betroffen sind, in etwa sowas wie ``je höher der Steuerfreibetrag, umso stärker der Bias''. Länder de erst sehr spät besteuern (und über nicht Besteuerte dann auch nicht Buch führen) haben einen krassen Bias. Wir könnten dann empfehlungen geben, ab welchem Perzentil man save interpretieren kann oder so.}

One can take a first look at the descriptives plotting the share of zeros over time seperately for each canton.

<<setup>>=
library(foreign)
df <- read.csv("../../data/ginis_und_perzentile_normal.csv",sep="\t")
df$zeros <- df$null_norm/df$cpop # i only use normal cases here
@

<<zero_descriptives>>=
library(ggplot2)
ggplot(df[df$steuerperiode>=1995,], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)+theme_bw()
ggplot(df[df$steuerperiode>=1995,], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)+geom_line(aes(y=G_steink),color="blue")+theme_bw()
@

@@

\begin{enumerate}
\item There is a small overall upward trend which we assume to be the Federal Administrations inflation adjustments to the tax threshold.
\item Geneva and Tessin show wild changes but those might be explained by the tax gap ("Bemessungslücke") that people exploited when the cantons changed the tax system. It remains unclear however why we can't see similar patter within other cantons.
\item There is some variance and we see different patterns over time and cantons. When estimating gini coefficients or the like we must therefore assume that ignoring ``the zeros'' leads to a bias that is not stable over time.
\end{enumerate}

@@

By adding the zeros as a seperate group we face several problems. With the exception of Geneva 1995/96 Pareto interpolation of percentiles p20 and above seems viable. Low percentiles however would need to be extrapolated so we might impose unrealistic assuptioms. Even when estimating ``safe'' measures above p20 (better p50) or a gini coefficient we need to make an assumption about the income structure of that group (a distribution, a mean income or zero-income). To be consistent with the measure of ``taxable income'' we could assume an income of zero for that group, calculate gini coefficients and compare them with the original (uncorrected/plain) version. To get a rough number we could (cantonwise) check the squared correlation between plain and corrected gini.

@@

The model outputs a test statisic for each canton that tells us whether the variation of the zero-rate over time leads to a significant deviation from the typical ``canton gini-level''. As the model has a decent fit we are not in great danger of omitted variable bias. Using a joint F-Test we can now test if all canton interactions are zero.

<<ftest, echo=FALSE>>=
library(survey, quietly=TRUE, warn.conflicts = FALSE)
ftest <- regTermTest(fit,"kanton:zeros")
@

In our case we can clearly reject the hypothesis that all interactions are zero ($p=\Sexpr{round(ftest$p,3)}$). This leads to the conclusion  that gini coefficients are biased by the variation of the zero-share which is kind of obvious but at the same time we can use the model to report adjusted gini coefficients. For example one might be interested in how inequality would had developed if the zero-share would have been constant over time. (Note RF: predict all data point using canton, time and the initial OR final zero-share to homogenize the time series)
Furthermore we can quantify how large the bias is and we can do this seperately for tax periods or seperately for cantons.

@@

What can we conclude from that analysis? First one must notice that aggregate measure like the gini (or others) do not always react in the same way when we cut off one part of the distribution, therefore the measures calculated from tax data is biased. On the other hand, the model coefficient of Switzerland as a whole is not significant suggesting that the cantonal biases cancel out each other. This seems plausible. Most of the ``tax optimization'' happens within Switzerland so the rich people who moved to Schwyz are now missing at another canton.

### 2014-05-27 paper/subdocuments/introduction.Rnw
fixed double quotes and R-paths so paper.Rnw compiles

% 1.Abschnitt: Bedeutung von Ungleichheit
Economic Resources can be seen as central indicator for life chances in general and a multitude of outcomes like physical and mental health, life expectancy and crime in particular (Wilkinson and Pickett 2009). While the study of social inequality can be considered as one of the core subjects of sociology in more recent years the concern about the widening gap was addressed by global leaders (WEF 2013) and scholars alike. Empirical evidence acknowledge the supposed trend that economic inequality increased in the majority of western countries over the last decades (OECD, 2008, 2011; Gornick and Jäntti, 2013;Salverda et al. 2014). Although the rise was not uniform, a common pattern seems to be identifiable, which can be referred to as the "hollowing of the the middle class" (Alderson and Doran 2013). Households are moving towards the top and the bottom of the distribution relative to the past, which is especially problematic as the middle class can be seen as the core of western democracies or as it is stated by Stiglitz (2012,117): "our democracy is being put at peril."
\\

@@

% 3.Abschnitt: What do we know about Switzerland
% Offical Data Collections on development of inequality
As we focus our paper on the case of Switzerland, it is important to embed our work in the context of given publications concering inequality in income. What is known about Switzerland so far? Looking for offical data, three main sources has to be mentioned, which can be considered as de facto offical data sources: EU-SILC, HBS and LIS-data. Figure XY shows the results steming from this three sources while looking at Gini of equivalised disposable income. Up to the day, EU-SILC or Statistics on Income and Living Conditions is the main source used for policy monitoring at EU-level. The main focus of EU-SILC is to collect data on a common "framework" to ensure comparability among EU-countries and countries living around or within the EU. As a Non-EU member Switzerland implemented the instrument not from the beginning (2004) but as from 2007. Therefore this times-series doesn't cover timepoints before 2007. As graph XY shows, following the results from SILC income inequality decreased from 2007 to 2013.\footnote{Data shown in the graph was downloaded from the Eurostat Metadata-portal http://epp.eurostat.ec.europa.eu/cache/ITY_SDDS/EN/ilc_esms.htm last accessed 21.Mai 2014.} The second important source concering the distribution of income is the Household Budget Survey (HBS). The main focus of this survey lays in providing detailed data on houshold budegts. This allows researcher to look at different income concepts like income before and after public transfers. Since 2000 the survey has been conducted on a continuous basis, which allows to look at a consistend time series from 2000 to 2011. As it can be seen from graph XY the trend is rather stable.\footnote{figures shown in the graph were calculated out of the original datasets, which were kindly provided by Swiss Federal Statistical Office.} Both time-series (SILC and HBS) cover a relatively short time period. A longer period is covered in the LIS-Data-set (1982-2004). Data-provider for the LIS Data is the Swiss Federal Statistical Office too. In contrast the the aformentioned surveys the LIS-data is harmonized out of three surveys: Swiss Income and Wealth Survey (1982), Swiss Poverty Survey (1992) and the Income and Consumption survey (2000,2002,2004).  All in all the LIS dataset contains the longest time series on inequality for Switzerland. Analyzing this data Gornick and Jäntti (2013) found for Switzerland a quite substantially decreases in income inequality, contradictory to the development in most other western countries. This result is supported by Grabka and Kuhn (2012) analyzing the Swiss Household Panel (2000-2009).\footnote{A further offical database for income distribution is the OECD-Database. It includes measures from Income and Consumption survey as well. Additional data for 2008 is available from EU-Survey of Income and Living Conditions (EU-SILC). But this change in survey is considered as a strict break. Comparison before and since 2008 is not recommended (OECD 2012:315). A further important database on inequality is the GINI database which has been derived from the GINI Country Reports (Nolan et al.,2014). But this dataset doesn't cover Switzerland.}  \\

<<Swiss inequality Plot>>=
library(ggplot2)
library(foreign)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
swissGini<-read.csv("data/swissGini.csv",header=TRUE,sep=";")
swissGini_long<-melt(swissGini,id="Year")
ggplot(data=swissGini_long,
  aes(y=value,x=Year),shape=variable)+
  ylab("Gini of equivalised disposable income") +
  scale_y_continuous(limits=c(0.18,0.4)) +
  scale_x_continuous(limits=c(1982,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=variable),size=3)+
  geom_line(data=swissGini_long[!is.na(swissGini_long$value),],aes(linetype=variable))+
  theme_bw()
@

@@

% 5. Abschnitt: Zwischenbilanz und Herausforderungen
Divergence can be explained with several factors. First of all, different data sources were used. The official data providers trust on survey data, whereas the later mentioned publications use tax data. It is assumed that the coverage of top incomes is better in tax data than it is in survey data (non-respondent bias), which is a crucial issue concerning inequality. On the other hand the focus on top income neglects other changes in the distribution of income as it is not possible to see, whether newer concerns like the "hollowing of the middle class" occurred in Switzerland or not, which leads to the second point. Different measure of inequality hampers the comparability. Third, different income concepts and different units of analysis were used. As it is shown by Modetta and Müller (2012) income distribution is strongly affected by governmental redistribution, reducing inequality substantially. With the focus on tax data the change in institutional settings is not covered. Also neglected is the household structure, whereas it is unclear how inequality is affected whether one looks at household income or at income of tax units. It can be assumed, that inequality corresponding to different concepts react differently on demographic change (change in household structure). \\

>>>>

% 1.Abschnitt: Bedeutung von Ungleichheit
Economic Resources can be seen as central indicator for life chances in general and a multitude of outcomes like physical and mental health, life expectancy and crime in particular (Wilkinson and Pickett 2009). While the study of social inequality can be considered as one of the core subjects of sociology in more recent years the concern about the widening gap was addressed by global leaders (WEF 2013) and scholars alike. Empirical evidence acknowledge the supposed trend that economic inequality increased in the majority of western countries over the last decades (OECD, 2008, 2011; Gornick and Jäntti, 2013;Salverda et al. 2014). Although the rise was not uniform, a common pattern seems to be identifiable, which can be referred to as the ``hollowing of the the middle class'' (Alderson and Doran 2013). Households are moving towards the top and the bottom of the distribution relative to the past, which is especially problematic as the middle class can be seen as the core of western democracies or as it is stated by Stiglitz (2012,117): ``our democracy is being put at peril.''
\\

@@

% 3.Abschnitt: What do we know about Switzerland
% Offical Data Collections on development of inequality
As we focus our paper on the case of Switzerland, it is important to embed our work in the context of given publications concering inequality in income. What is known about Switzerland so far? Looking for offical data, three main sources has to be mentioned, which can be considered as de facto offical data sources: EU-SILC, HBS and LIS-data. Figure XY shows the results steming from this three sources while looking at Gini of equivalised disposable income. Up to the day, EU-SILC or Statistics on Income and Living Conditions is the main source used for policy monitoring at EU-level. The main focus of EU-SILC is to collect data on a common ``framework'' to ensure comparability among EU-countries and countries living around or within the EU. As a Non-EU member Switzerland implemented the instrument not from the beginning (2004) but as from 2007. Therefore this times-series doesn't cover timepoints before 2007. As graph XY shows, following the results from SILC income inequality decreased from 2007 to 2013.\footnote{Data shown in the graph was downloaded from the Eurostat Metadata-portal http://epp.eurostat.ec.europa.eu/cache/ITY_SDDS/EN/ilc_esms.htm last accessed 21.Mai 2014.} The second important source concering the distribution of income is the Household Budget Survey (HBS). The main focus of this survey lays in providing detailed data on houshold budegts. This allows researcher to look at different income concepts like income before and after public transfers. Since 2000 the survey has been conducted on a continuous basis, which allows to look at a consistend time series from 2000 to 2011. As it can be seen from graph XY the trend is rather stable.\footnote{figures shown in the graph were calculated out of the original datasets, which were kindly provided by Swiss Federal Statistical Office.} Both time-series (SILC and HBS) cover a relatively short time period. A longer period is covered in the LIS-Data-set (1982-2004). Data-provider for the LIS Data is the Swiss Federal Statistical Office too. In contrast the the aformentioned surveys the LIS-data is harmonized out of three surveys: Swiss Income and Wealth Survey (1982), Swiss Poverty Survey (1992) and the Income and Consumption survey (2000,2002,2004).  All in all the LIS dataset contains the longest time series on inequality for Switzerland. Analyzing this data Gornick and Jäntti (2013) found for Switzerland a quite substantially decreases in income inequality, contradictory to the development in most other western countries. This result is supported by Grabka and Kuhn (2012) analyzing the Swiss Household Panel (2000-2009).\footnote{A further offical database for income distribution is the OECD-Database. It includes measures from Income and Consumption survey as well. Additional data for 2008 is available from EU-Survey of Income and Living Conditions (EU-SILC). But this change in survey is considered as a strict break. Comparison before and since 2008 is not recommended (OECD 2012:315). A further important database on inequality is the GINI database which has been derived from the GINI Country Reports (Nolan et al.,2014). But this dataset doesn't cover Switzerland.}  \\

<<Swiss inequality Plot>>=
library(ggplot2)
library(foreign)
library(reshape, quietly=TRUE, warn.conflicts = FALSE)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
swissGini<-read.csv("../../data/swissGini.csv",header=TRUE,sep=";")
swissGini_long<-melt(swissGini,id="Year")
ggplot(data=swissGini_long,
  aes(y=value,x=Year),shape=variable)+
  ylab("Gini of equivalised disposable income") +
  scale_y_continuous(limits=c(0.18,0.4)) +
  scale_x_continuous(limits=c(1982,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=variable),size=3)+
  geom_line(data=swissGini_long[!is.na(swissGini_long$value),],aes(linetype=variable))+
  theme_bw()
@

@@

% 5. Abschnitt: Zwischenbilanz und Herausforderungen
Divergence can be explained with several factors. First of all, different data sources were used. The official data providers trust on survey data, whereas the later mentioned publications use tax data. It is assumed that the coverage of top incomes is better in tax data than it is in survey data (non-respondent bias), which is a crucial issue concerning inequality. On the other hand the focus on top income neglects other changes in the distribution of income as it is not possible to see, whether newer concerns like the ``hollowing of the middle class'' occurred in Switzerland or not, which leads to the second point. Different measure of inequality hampers the comparability. Third, different income concepts and different units of analysis were used. As it is shown by Modetta and Müller (2012) income distribution is strongly affected by governmental redistribution, reducing inequality substantially. With the focus on tax data the change in institutional settings is not covered. Also neglected is the household structure, whereas it is unclear how inequality is affected whether one looks at household income or at income of tax units. It can be assumed, that inequality corresponding to different concepts react differently on demographic change (change in household structure).

### 2014-05-27 paper/subdocuments/results.Rnw
update minor changes
fixed double quotes and R-paths so paper.Rnw compiles

<<setup>>=
library(foreign)
library(ggplot2)
df <- read.dta("data/ginis_und_perzentile.dta")
df$zeros <- df$null_norm/df$cpop # i only use normal cases here
@

@@

<<with_without_zeros>>=
dflong <- melt(df[,c("G_steink","G_steink0","kanton","steuerperiode")], id=c("kanton","steuerperiode"))
names(dflong)[3]<-"Type"
names(dflong)[4]<-"Gini"
levels(dflong$Type)<-c("without zeros","including zeros")
ggplot(dflong[dflong$kanton=="CH" & dflong$steuerperiode>=1995.5,], aes(x=steuerperiode,y=Gini,colour=Type))+geom_line(size=1)+theme_bw()+xlab("tax period")
ggplot(dflong[dflong$steuerperiode>=1995.5,], aes(x=steuerperiode,y=Gini,colour=Type))+geom_line()+facet_wrap(~kanton)+theme_bw()+xlab("tax period")+ opts(axis.text.x=theme_text(angle=60))

>>>>

<<setup_gini_data>>=
library(foreign)
library(ggplot2)
df <- read.csv("../../data/ginis_und_perzentile_normal.csv",sep="\t")
df$zeros <- df$null_norm/df$cpop # i only use normal cases here
@

@@

<<with_without_zeros,warning=FALSE>>=
dflong <- melt(df[,c("G_steink","G_steink0","kanton","steuerperiode")], id=c("kanton","steuerperiode"))
names(dflong)[3]<-"Type"
names(dflong)[4]<-"Gini"
levels(dflong$Type)<-c("without zeros","including zeros")
ggplot(dflong[dflong$kanton=="CH" & dflong$steuerperiode>=1995.5,], aes(x=steuerperiode,y=Gini,colour=Type))+geom_line(size=1)+theme_bw()+xlab("tax period")
ggplot(dflong[dflong$steuerperiode>=1995.5,], aes(x=steuerperiode,y=Gini,colour=Type))+geom_line()+facet_wrap(~kanton)+theme_bw()+xlab("tax period")+theme(axis.text.x = element_text(angle = 60, hjust = 1))
